---
title: "Recovering Spatial Ecotypes from Single-Cell Spatial Transcriptomics Data"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

In this tutorial, we will illustrate how to recover spatial ecotypes (SEs) from single-cell-scale spatial transcriptomics (ST) data, generated by platforms such as MERSCOPE, Xenium, CosMx, and Visium HD.

__First load required packages for this vignette__

```{r lib}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(parallel))
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(data.table))
library(SpatialEcoTyper)
```

# Data preparation

SE recovery for single-cell ST data requires two inputs:

* __Gene expression matrix__: a numeric matrix with genes as rows and cells as columns.
* __Metadata__: a data frame containing at least three columns: "X" (x-coordinate), "Y" (y-coordinate), and "CellType" (cell type annotation). Row names must match the column names (cell IDs) of the expression matrix.

<details open><summary><strong>Text files as input</strong></summary>

```{r loaddtsv}
# Load metadata
scmeta <- read.table("https://spatialecotyper.stanford.edu/inc/inc.public.vignettes.php?file=CRC2_subset_scmeta.tsv",
                      sep = "\t", header = TRUE, row.names = 1)
head(scmeta[, c("X", "Y", "CellType")])

# Load expression matrix
scdata <- fread("https://spatialecotyper.stanford.edu/inc/inc.public.vignettes.php?file=CRC2_subset_counts.tsv.gz",
                sep = "\t",header = TRUE, data.table = FALSE)
rownames(scdata) <- scdata[, 1]
scdata <- as.matrix(scdata[, -1])
head(scdata[,1:5])

# Ensure matching cell order
scdata <- scdata[, match(rownames(scmeta), colnames(scdata))]
```
</details>

## Data normalization

The data should be normalized to a log-transformed CPM (counts per million) or a similar scale. The `NormalizeData` function is applied at this step to perform this normalization.

```{r normdat}
normdata = NormalizeData(scdata)
```

# SE recovery

The [RecoverSE](../reference/RecoverSE.html) function will be used to assign single cells into SEs. Users can either use default model to recover predefined SEs or use custom model to recover newly defined SEs.

**Note**: Cell type annotation (`celltypes` argument) is required.

<details open><summary><strong>Using default models</strong></summary>

The default NMF models were trained on MERSCOPE data across carcinomas and melanoma. The model supports the following cell types: B cells, CD4+ T cells, CD8+ T cells, NK cells, plasma cells, macrophages, dendritic cells, fibroblasts, and endothelial cells.
All cells in the query data should be grouped into one of "B", "CD4T", "CD8T", "NK", "Plasma", "Macrophage", "DC", "Fibroblast", and "Endothelial", case sensitive. All the other cell types will be assigned to NonSE. 

```{r scrna_recov}
sepreds <- RecoverSE(normdata, 
                     celltypes = scmeta$CellType,
                     ncores = 8)
head(sepreds)
```

The SE recovery output contains five columns: cell ID (CID), cell type (CellType), initial SE assignment (InitSE), final SE assignment (SE), and prediction score (PredScore).

</details>

<details><summary><strong>Using custom models</strong></summary>

To use custom model, users should first develop a model following the tutorial [NMF Model Development for Spatial Ecotype Recovery from Single-Cell and Spatial Transcriptomics Data](TrainRecoveryModels.html). The resulting model can be used for SE recovery. An example model is available at [SE_Recovery_W_list.rds](https://spatialecotyper.stanford.edu/inc/inc.public.vignettes.php?file=SE_Recovery_W_list.rds).

Download a custom model
```{r downloadWs, eval=FALSE}
url <- "https://spatialecotyper.stanford.edu/inc/inc.public.vignettes.php?file=SE_Recovery_W_list.rds"
download.file(url, destfile = "SE_Recovery_W_list.rds", mode = "wb")
```

Load the custom model
```{r viewWs, eval=FALSE}
Ws <- readRDS("SE_Recovery_W_list.rds")
names(Ws) ## named list of W matrices
head(Ws[[1]]) ## feature by SE matrix
```

Using custom model for SE recovery by specifying the `Ws` argument.
```{r scrna_recov2, eval=FALSE}
sepreds <- RecoverSE(normdata, celltypes = scmeta$CellType, Ws = Ws)
```

After prediction, we recommend retaining only assignments corresponding to SE-specific cell states identified in [Tutorial 3](https://digitalcytometry.github.io/spatialecotyper/articles/Discovery_SE_CellStates.html). Cells not assigned to these SE-specific states should be grouped as "NonSE," as they cannot be robustly mapped to corresponding SE. 

The SE recovery output contains five columns: cell ID (CID), cell type (CellType), initial SE assignment (InitSE), final SE assignment (SE), and prediction score (PredScore). By default, assignments with PredScore ≥ 0.6 are retained as high-confidence results. For custom models, users are encouraged to determine the optimal threshold based on leave-one-sample-out cross-validation performance.

</details>

__Combine SE recovery results into metadata__

```{r updatemeta, eval=TRUE}
# Add SE assignments into the metadata
scmeta$SE = sepreds$SE[match(rownames(scmeta), rownames(sepreds))]
```

# Validation: spatial colocalization of SE cell states

After recovering SE-specific cell states from single-cell ST data, users can validate the SEs by testing whether cell states belonging to the same SE are more spatially colocalized than expected by random chance using the [Colocalization](../reference/Colocalization.html) function. The spatial coordinates (column X and Y), cell type (CellType) and cell state annotations (State) are required for the analysis.

**Note**: By default (`test = TRUE`), the function evaluates the significance of cell state co-localization within each SE and returns a vector of p-values. This analysis requires cell state names to follow the format “SE1_CD4T”, where an underscore separates the SE label from the cell type, enabling extraction of SE groupings from the cell state names.

```{r coloc, eval=TRUE}
# Verify the four columns required for the analysis.
head(scmeta[, c("X", "Y", "SE", "CellType")])

## For quick test run, 100 permutations were performed to compute the colocalization scores.
coloc_res = Colocalization(scmeta, coords = c("X", "Y"), 
                           SE = "SE", CellType = "CellType",
                           radius = 50, nperm = 100, 
                           test = TRUE, ncores = 8)
head(coloc_res$ColocIndex[, 1:5]) ## Matrix of colocalization indices between all cell states
coloc_res$Pval ## Vector of p-values for colocalization significance
```

You can visualize the results using the [CooccurrenceHeatmapView](../reference/CooccurrenceHeatmapView.html) function.
```{r visualcoloc}
p = CooccurrenceHeatmapView(coloc_res$ColocIndex, coloc_res$Pval)
p = ComplexHeatmap::draw(p)
ses = gsub("_.*", "", rownames(coloc_res$ColocIndex))
drawRectangleAnnotation(p, ses, ses)
```

Here, in this small demo data, we can see strong colocalization of cell states within most SEs, whereas SE9 shows a weaker pattern, although it approaches statistical significance (P < 0.1). In contrast, the NonSE group, serving as a negative control, does not exhibit spatial colocalization.

# Validation: spatial autocorrelation

Users can also validate the SEs by assessing the spatial autocorrelation with Moran’s I. To control for bias, permutation experiments will be performed to normalize Moran’s I into z-scores using the [ComputeNormalizedMoranI](../reference/ComputeNormalizedMoranI.html) function.

```{r morani, eval=TRUE}
## For quick test run, test with 10 permutations were performed.
morani_zscore = ComputeNormalizedMoranI(scmeta, coords = c("X", "Y"), 
                                        SE = "SE", CellType = "CellType", 
                                        nperm = 10, ncores = 8)
morani_zscore
```


# Validation: average expression of SE markers

Users can also validate the SEs by visualizing the average expression of SE consensus markers across recovered SEs using the [AverageMarkerExpression](../reference/AverageMarkerExpression.html) function.

```{r visualconsensus}
## The function is built on Seurat functions, so first create a Seurat object.
obj = CreateSeuratObject(scdata, meta.data = scmeta)
obj = NormalizeData(obj)
```

By default, the [AverageMarkerExpression](../reference/AverageMarkerExpression.html) function compute the average expression of SE consensus markers from [Zhang et al., Nature 2026 paper](https://www.nature.com/articles/s41586-026-10452-4).

```{r visualconsensus2}
## Visualize average expression of SE consensus markers across recovered SEs
res = AverageMarkerExpression(obj, group.by = "SE")
plot(res$p)

# Underlying data
res$AvgExp
```

The [AverageMarkerExpression](../reference/AverageMarkerExpression.html) function can also be used to visualize the average expression of user-defined gene sets by specifying the `markers` parameter as a list of gene lists.

```{r othermarkers}
genesets = list(set1 = c("FOS", "EGR1", "DUSP1", "JUNB", "JUN", "FOSB"),
                set2 = c("SPP1", "CCL2", "MMP9"),
                set3 = c("CXCL12", "TGFBR2", "PDK4", "HLA-DMA"),
                set4 = c("ACTA2", "CAV1", "CLDN5"),
                set5 = c("LRP1", "MAFB"),
                set6 = c("CXCL2", "NFKBIA", "CDKN1A"),
                set7 = c("STAT1", "TAP1", "HLA-A", "HLA-B", "HLA-C", "B2M"),
                set8 = c("PKM", "PCNA", "BST2"),
                set9 = c("NRP1", "ITGB1", "CD276"))
res = AverageMarkerExpression(obj, group.by = "SE", genesets = genesets)
plot(res$p)
```

# Session info

The session info allows users to replicate the exact environment and identify potential discrepancies in package versions or configurations that might be causing problems.

```{r session}
sessionInfo()
```
