---
title: "Inferring Spatial Ecotype Abundances from Bulk Gene Expression Data"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

In this tutorial, we will demonstrate how to infer spatial ecotype (SE) abundances from bulk gene expression profiles, including bulk RNA-seq and Visium spatial transcriptomics. 

__First load required packages for this vignette__

```{r lib}
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(data.table))
library(SpatialEcoTyper)
```


# Inferring SE levels from bulk RNA-seq data

We will start with the SE deconvolution from bulk RNA-seq. We will use bulk RNA-seq data from TCGA melanoma (SKCM) samples for demonstration. The gene expression matrix is available at [SKCM_RNASeqV2.geneExp.tsv](https://spatialecotyper.stanford.edu/inc/inc.public.vignettes.php?file=SKCM_RNASeqV2.geneExp.tsv), which is a Transcripts Per Million (TPM) matrix obtained from the [PanCanAtlas](https://gdc.cancer.gov/about-data/publications/pancanatlas).

## Loading bulk expression of TCGA melanoma samples

```{r loadbulk}
bulkdata <- fread("https://spatialecotyper.stanford.edu/inc/inc.public.vignettes.php?file=SKCM_RNASeqV2.geneExp.tsv", sep = "\t", data.table = FALSE)
rownames(bulkdata) <- bulkdata[, 1] ## Set the first column as row names
bulkdata <- bulkdata[, -1] ## Drop the first column
head(bulkdata[, 1:5]) ## TPM matrix
```

## SE deconvolution

The [DeconvoluteSE](../reference/DeconvoluteSE.html) function infers SE abundances in bulk tissue samples. Users can either use the default model, which estimates the abundances of predefined SEs, or apply a custom model to infer the abundance of newly defined SEs from bulk gene expression data. Before prediction, [DeconvoluteSE](../reference/DeconvoluteSE.html) automatically applies a log₂ transformation (if the maximum value exceeds 80) and standardizes the data so that each gene has a mean of 0 and a unit variance. To ensure reliable standardization, we recommend a minimum of 20 samples. For smaller datasets, we recommend to combine them with a larger dataset for normalization prior to running [DeconvoluteSE](../reference/DeconvoluteSE.html).

The default model is trained on pseudobulk gene expression data and predicts the abundances of nine predefined SEs along with a nonSE group, which includes cancer cells and cells not associated with any SEs. The predicted SE and nonSE abundances for each sample sum to 1, making them comparable across samples.

### Using default model

```{r DeconvoluteSE1}
sefracs <- DeconvoluteSE(bulkdata, scale = TRUE)
head(sefracs)
```
**Note**: If your data are already standardized to have zero mean and unit variance per gene, set `scale = FALSE`.

### Using custom model

After identifying SEs using [SpatialEcoTyper](../reference/SpatialEcoTyper.html) or [MultiSpatialEcoTyper](../reference/MultiSpatialEcoTyper.html), users can develop a Non-Negative Matrix Factorization (NMF) model to deconvolve SEs from bulk tissue samples, following the [Tutorial 7](TrainDeconvModel.html).

__Deconvolution__

Users can train an NMF model for SE deconvolution following the tutorial above. After that, the new model can be used for SE deconvolution by specifying the `W` parameter:
```{r DeconvoluteSE2, eval=FALSE}
# Do not run; replace new_model
sefracs <- DeconvoluteSE(bulkdata, W = new_model)
```


## Visualizing SE abundances across samples
```{r visualse}
library(grid)
HeatmapView(sefracs, breaks = c(0, 0.15, 0.3), 
            show_row_names = FALSE, cluster_rows = TRUE)
```


# Inferring SE levels from Visium data

Here, we will demonstrate how to infer SE levels in Visium spots using a breast cancer sample. The expression data can be accessed from: [Visium_Breast_Cancer_filtered_feature_bc_matrix.h5](https://spatialecotyper.stanford.edu/inc/inc.public.vignettes.php?file=Visium_Breast_Cancer_filtered_feature_bc_matrix.h5), which was obtained from [10x Genomics datasets](https://www.10xgenomics.com/datasets).

## Loading data

First, download and load the expression data into R.
```{r loadvisium}
if(!"hdf5r" %in% installed.packages()) BiocManager::install("hdf5r")
require("hdf5r")
url <- "https://spatialecotyper.stanford.edu/inc/inc.public.vignettes.php?file=Visium_Breast_Cancer_filtered_feature_bc_matrix.h5"
download.file(url, destfile = "Visium_Breast_Cancer_filtered_feature_bc_matrix.h5", mode = "wb")
# Load Visium gene expression matrix. Rows are genes, columns are spots.
visiumdat <- Read10X_h5("Visium_Breast_Cancer_filtered_feature_bc_matrix.h5")
# normalize the expression data
visiumdat <- NormalizeData(visiumdat)

meta <- read.csv("https://spatialecotyper.stanford.edu/inc/inc.public.vignettes.php?file=Visium_Breast_Cancer_tissue_positions_list.csv", 
                 header = FALSE, row.names = 1)
colnames(meta) <- c("tissue", "row", "col", "imagerow", "imagecol")
meta <- meta[colnames(visiumdat), ]
head(meta)
```

## SE deconvolution

The process of SE deconvolution for Visium data is the same as for bulk RNA-seq data described above, using the [DeconvoluteSE](../reference/DeconvoluteSE.html) function.

```{r DeconvoluteSE3}
sefracs <- DeconvoluteSE(visiumdat, scale = TRUE)
head(sefracs)
```


## Visualizing SE abundances across Visium spots

The [SpatialView](../reference/SpatialView.html) function can be used to visualize SE levels across Visium spots.

```{r DeconvoluteSE4}
library(ggplot2)
## attach the spatial coordinates to the matrix for visualization
gg = cbind(sefracs, meta[match(rownames(sefracs), rownames(meta)), ])
gg = as.data.frame(gg)
SpatialView(gg, by = "SE7", X = "imagerow", Y = "imagecol", pt.size = 2) +
  scale_color_gradientn(colors = rev(RColorBrewer::brewer.pal(11, "Spectral"))) + 
  coord_fixed()
```

## Aggregating SE abundances across Visium spots

To compute sample-level SE abundances, Spatial EcoTyper calculates a weighted average of SE abundances across all Visium spots, where each spot is weighted by its inferred number of cells. The resulting SE abundances are then normalized to sum to 1.

```{r samplelevel}
ncells = InferNCells(visiumdat)

se_abundance = colSums(sefracs * ncells)
se_abundance = se_abundance / sum(se_abundance)

se_abundance
```


# Session info

The session info allows users to replicate the exact environment and identify potential discrepancies in package versions or configurations that might be causing problems.

```{r session}
sessionInfo()
```
